---
title: "Homework 3, P8130"
author: "Emil Hafeez (eh2928)"
date: "10/15/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r results 'hide', warning = FALSE}
library(tidyverse)

exercise_df =
  read_csv(
      "./data/exercise.csv")
```

# Problem 1

For each question, make sure to state the formulae for hypotheses, test-statistics, decision rules/p-values, and provide interpretations in the context of the problem. Use a type I error of 0.05 for all tests.

### Problem 1.a

Perform appropriate tests to assess if the Systolic BP at 6 months is significantly different from the baseline values for the intervention group.

********Check for normality, consider a bonferroni adjustment, state that it's a a paired two-tailed test using an estimate. Then provide the formulas for the test statistic needed, critical value needed, the value, and then get these. and then interpret in context.********

In order to determine an appropriate test, we first check for normality using visual inspection of the plot, using raw data to determine $p$ and $1-p$. from a simple heuristic of $np(n-p) \geq 5$. In this case, n = 36, p = 0.05, q = 0.95, and so 


#### Problem 1.a.i


#### Problem 1.a.ii





#### Problem 1.b




### Problem 1.c

#### Problem 1.c.i


#### Problem 1.c.ii






